---
title: "Project Report MovieLens"
author: "Samuel Kuenti"
date: "2. April 2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction
*An introduction/overview/executive summary section that describes the data set and summarizes the goal of the project and key steps that were performed*

The task for this first project of the Capstone module was to build a machine learning algorithm.  
The idea was to predict movie ratings based on a set of features, with the model being trained on
a large part of the data and then evaluated by means of a holdout set.

The data came from the publicly available MovieLens-databse. The data were split into a training set,
a test set and the validation set. The latter was only used after the development of the model in order to assess its performance.

In the present solution, in addition to the movie and user effects already modeled in the 
Machine Learning module, a small genre effect was established, which was used to refine the model.

This refined model was assessed by means of the validation set of the original data, which mimicked model performance on actual 'real-world' data.

## Methods
methods/analysis section that explains the process and techniques used, including data cleaning, data exploration and visualization, insights gained, and your modeling approach

The modeling data were split 90/10% into a training and a test set. This splitting was chosen because the model's structure was clear from the beginning, and no tweaking was expected in response to model performance.

The training set consisted of 'r edx_train %>% summarise(n=n())' cases.



# Is there a genere effect?
If yes, I which could rely on to refine predictions.

It tunr out that genre data is difficult, because each movie is associated to multiple
genres.

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

RMSE is defined as follows: 

RMSE = sqrt(1/N*sum((y_hat_u_i-y_u_i)^2)

## Results
a results section that presents the modeling results and discusses the model performance

## Conclusion
a conclusion section that gives a brief summary of the report,its limitations and future work
