---
title: "Project Report MovieLens"
author: "Samuel Kuenti"
date: "2 4 2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In order to create a movie recommendation system, a machine learning algorithm had to 
be trained on a part of the MovieLens dataset. The idea was to predict movie ratings 
based on a set of features.

Here, in addition to the movie and user effects already modelled in the course, it was
tested whether there was a genre effect, and if yes, this genre effect was added to the model
in addition to the other two effects (movie and user) in order to improve predictions.

The model achieving the lowest residual mean squared error (RMSE) after training 
was then chosen and used to assess the final model 
by means of an exclusive holdout set of the original data, which mimicked model performance
on actual 'real-world' data.

Maybe this will be an attempt at improving the model by including genre data

```{r cars}
summary(cars)
```

## Methods

Data was split 80/20% into a training and test set in order to tune an algorithm.

# Is there a genere effect?
If yes, I which could rely on to refine predictions.

It tunr out that genre data is difficult, because each movie is associated to multiple
genres.

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

RMSE is defined as follows: 

RMSE = sqrt(1/N*sum((y_hat_u_i-y_u_i)^2)

## Results


## Conclusion
